Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-09-10 journal paper/content/sec-bg.tex
Minor edits.

We present self-contained background on multi-armed bandits (MAB), to make the paper accessible to researchers who are not experts on MAB. We focus on three algorithm classes, as in Section~\ref{sec:sim}:

@@

We list the precise upper and lower bounds on the regret rates only to illustrate how the three algorithm classes are separated from one another; the exact results are not essential for the paper. We use standard conventions: $O(f(t))$ and $\Omega(f(t))$ means at most (resp., at least) $f(n)$, up to constant factors, starting from large enough $t$. Likewise, $\tildeO(f(t))$ notation suppresses the $\polylog(t)$ factors.

@@

The optimal regret rates are achieved by several adaptive-exploration  algorithms, of which the most known are
Thompson Sampling \citep{Thompson-1933,TS-survey-FTML18},%
\footnote{While Thompson Sampling is the first and arguably best-known bandit algorithm, the corresponding regret bounds have been proved relatively recently \citep{Shipra-colt12,Kaufmann-alt12,Shipra-aistats13}.}
UCB1 \citep{bandits-ucb1},
and Successive Elimination \citep{EvenDar-icml06}. There is a substantial amount of follow-up work on refining these regret rates, but it is not as relevant to this paper.

Exploration-separating algorithms completely separate exploration and exploitation. Ahead of time, each round is either selected for exploration, in which case  the distribution over arms does not depend on the observed data, or it is assigned to exploitation, in which case the data from this round is discarded. The simplest approach, called \emph{Explore-First}, explores uniformly for a predetermined number of rounds, then chooses one arm for ``exploitation" and uses it from then on. A more refined approach, called \emph{Epsilon-Greedy}, explores uniformly in each round with a predetermined probability, and ``exploits" with the remaining probability. Both algorithms, and the associated $\tildeO(T^{2/3})$ regret bounds, have been ``folklore knowledge" for decades. The general definition of ``exploration-separating algorithms" and the associated lower bounds trace back to \citet{MechMAB-ec09}.%
\footnote{\citet{MechMAB-ec09} consider a closely related, but technically different setting, which can be easily ``translated" into ours (either as a corollary or as another application of the same proof technique). Detailing this is beyond the scope of this appendix.}

@@

\xhdr{Advanced aspects.}
Switching from ``greedy" to ``exploration-separating" to ``adaptive-exploration" algorithms involves substantial adoption costs in infrastructure and personnel training \citep{DS-arxiv}. Inserting exploration into a complex decision-making pipeline necessitates a substantial awareness of the technology and a certain change in a mindset for the personnel, as well as an infrastructure to collect and analyze the data. Adaptive exploration requires the said infrastructure to propagate the data analysis back to the ``front-end" where the decisions are made, and do it on a sufficiently fast and regular cadence. Framing the problem (in terms of the appropriate modeling assumptions and action features) and debugging the machine learning code tend to be quite subtle, too.

The lower bounds mentioned above are fairly typical, in the sense that they hold for a wide variety of problem instances, even though they are usually (and most cleanly) presented as worst-case. The $\Omega(\sqrt{T})$ lower bound from \citet{bandits-exp3} can be extended to hold for most problem instances, in the following sense: for each instance $\mI$ there exists a ``decoy instance" $\mI'$ such that any algorithm incurs regret $\Omega(\sqrt{T})$ on at least one of them. The ``gap-dependent" lower bound of
 $\Omega(\tfrac{K}{\Delta}\log T)$
in fact holds for all problem instances and all algorithms that are not \emph{terrible} on the large-gap instances \citep{Lai-Robbins-85}. The $\Omega(T^{2/3})$ lower bound for exploration-separating algorithms in fact applies to all problem instances, as long as the algorithm achieves $\tildeO(T^{2/3})$ regret rate in the worst case \citep{MechMAB-ec09}.
Moreover, there is a tradeoff between the worst-case upper bound on the regret rate and a lower bound that applies for all problem instances \citep[Theorem 4.3 in][using the same proof technique]{MechMAB-ec09}.

Some MAB algorithms are Bayesian, \eg Thompson Sampling,  in the sense that they input a prior on mean rewards, and attain strong Bayesian guarantees (in expectation over the prior) when the prior is correct. Such algorithms can also be initialized with some simple `fake' priors; in fact, Thompson Sampling satisfies the optimal regret bounds listed above.

@@

The three classes are well-separated in terms of \BIR, much like they are in terms of regret.
%(The precise regret rates are listed below only to illustrate this point.)
\begin{itemize}
\item \DynGreedy has at least a constant \BIR for many reasonable priors (where the constant can depend on $K$ and the prior, but not on $t$). The reason and the proof are exactly the same as for regret.

>>>>

We present self-contained background on multi-armed bandits (MAB), to make the paper accessible to researchers who are not experts on MAB. More details can be found in any of the monographs
\citep{Bubeck-survey12,slivkins-MABbook,LS19bandit-book}.

We focus on three algorithm classes, as in Section~\ref{sec:sim}:

@@

While we list precise upper and lower bounds on the regret rates, the main goal is to illustrate how the three algorithm classes are separated from one another; the exact results are not essential for this paper. For ease of presentation, we use standard asymptotic notation from computer science: $O(f(t))$ and $\Omega(f(t))$ means at most (resp., at least) $f(n)$, up to constant factors, starting from large enough $t$. Likewise, $\tildeO(f(t))$ notation suppresses the $\polylog(t)$ factors.

@@

The optimal regret rates are achieved by several adaptive-exploration  algorithms, of which the most known are
Thompson Sampling \citep{Thompson-1933,TS-survey-FTML18},%
\footnote{While Thompson Sampling dates back to 1933 and is probably the best-known bandit algorithm, its regret has not been understood until recently \citep{Shipra-colt12,Kaufmann-alt12,Shipra-aistats13}.}
UCB1 \citep{bandits-ucb1},
and Successive Elimination \citep{EvenDar-icml06}.%
\footnote{A substantial follow-up work on more ``refined" regret rates is not as relevant to this paper.}
These algorithms are very simple to describe. Focus on one round and consider the posterior distribution and/or the confidence interval on each arm's mean reward. Thompson Sampling draws a sample (``score") from each arm's posterior distribution, and picks an arm with the largest score. UCB1 picks an arm with the largest upper confidence bound. Successive Elimination eliminates an arm once it is worse than some other arm with high confidence, and chooses uniformly among the remaining arms.

Exploration-separating algorithms completely separate exploration and exploitation. Ahead of time, each round is either selected for exploration, in which case  the distribution over arms does not depend on the observed data, or it is assigned to exploitation, in which case the data from this round is discarded. The simplest approach, called \emph{Explore-First}, explores uniformly for a predetermined number of rounds, then chooses one arm for ``exploitation" and uses it from then on. A more refined approach, called \emph{Epsilon-Greedy}, explores uniformly in each round with a predetermined probability, and ``exploits" with the remaining probability. Both algorithms, and the associated $\tildeO(T^{2/3})$ regret bounds, have been ``folklore knowledge" for decades. The general definition and lower bounds trace back to \citet{MechMAB-ec09}.%
\footnote{\citet{MechMAB-ec09} consider a closely related, but technically different setting, which can be easily ``translated" into ours (either as a corollary or as another application of the same proof technique).}

@@

\xhdr{Advanced aspects.}
Switching from ``greedy" to ``exploration-separating" to ``adaptive-exploration" algorithms involves substantial adoption costs in infrastructure and personnel training \citep{DS-arxiv}. Inserting exploration into a complex decision-making pipeline necessitates a substantial awareness of the technology and a certain change in mindset, as well as an infrastructure to collect and analyze the data. Adaptive exploration requires the said infrastructure to propagate the data analysis back to the ``front-end" where the decisions are made, and do it on a sufficiently fast and regular cadence. Framing the problem (\eg choosing modeling assumptions and action features) and debugging the machine learning algorithms tend to be quite subtle, too.

The lower bounds mentioned above are fairly typical: while they are usually (and most cleanly) presented as worst-case, the actually hold for a wide variety of problem instances. The $\Omega(\sqrt{T})$ lower bound from \citet{bandits-exp3} can be extended to hold for most problem instances, in the following sense: for each instance $\mI$ there exists a ``decoy instance" $\mI'$ such that any algorithm incurs regret $\Omega(\sqrt{T})$ on at least one of them. The ``gap-dependent" lower bound of
 $\Omega(\tfrac{K}{\Delta}\log T)$
in fact holds for all problem instances and all algorithms that are not \emph{terrible} on the large-gap instances \citep{Lai-Robbins-85}. The $\Omega(T^{2/3})$ lower bound for exploration-separating algorithms in fact applies to all problem instances, as long as the algorithm achieves $\tildeO(T^{2/3})$ regret rate in the worst case \citep{MechMAB-ec09}.%
\footnote{Moreover, there is a tradeoff between the worst-case upper bound on the regret rate and a lower bound that applies for all problem instances \citep[Theorem 4.3 in][]{MechMAB-ec09}.}

Some MAB algorithms, \eg Thompson Sampling, are Bayesian: they input a prior on mean rewards, and attain strong Bayesian guarantees (in expectation over the prior) when the prior is correct. Such algorithms can also be initialized with some simple `fake' priors; in fact, this is how Thompson Sampling can be made to satisfy the optimal regret bounds listed above.

@@

The three classes are well-separated in terms of \BIR, much like they are in terms of regret.
%(The precise regret rates are listed below only to illustrate this point.)
\begin{itemize}
\item \DynGreedy has at least a constant \BIR for many reasonable priors (where the constant can depend on $K$ and the prior, but not on $t$). The reason / proof is the same as for regret.
